---
title: "Pivot Types: Reshaping vs Cross-Tabulation"
format:
  html:
    code-fold: true
    code-summary: "Show the code"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

## Introduction

The word "pivot" covers two different operations:

- **Pure reshaping** — changing the *shape* of data (long → wide) without performing any arithmetic on values. Data is redistributed into new columns; nothing is summarized or collapsed.
- **Cross-tabulation** — reshaping *plus* aggregation. Multiple source rows sharing the same (row-group, column-group) combination are collapsed into a single cell via a summary function (mean, count, sum, etc.).

We'll demonstrate both using the `gapminder` dataset, which records country-level statistics at five-year intervals from 1952 to 2007. The variables we'll use are:

- `country` — the unit of observation
- `continent` — the geographic grouping
- `year` — the time point (1952, 1957, ..., 2007)
- `lifeExp` — life expectancy at birth (years)

```{r}
library(gapminder)
library(tidyverse)
library(DBI)
library(duckdb)
library(RPostgres)

data(gapminder)
glimpse(gapminder)
```

---

## Part 1: Pure Reshaping

### The data structure

For pure reshaping, the requirement is that each combination of row-identifier and spreading-column value appears **exactly once** in the source data. Thus there is nothing to aggregate, only redistribute.

In our first example:

- **Rows:** country
- **Columns:** year (each distinct year becomes a column header)
- **Cells:** life expectancy

Each country has exactly one life expectancy per year, so no aggregation is needed or possible. This is the defining condition for pure reshaping.

```{r}
# Confirm each country-year combination is unique
gapminder |>
  count(country, year) |>
  filter(n > 1)
```

No rows — every country-year combination is unique. Pure reshaping is appropriate here.

### Approach 1: `pivot_wider()` in R (tidyverse)

`pivot_wider()` is the canonical R function for pure reshaping. It takes three arguments that map directly to the three pivot roles:

- `id_cols` — the row-grouping specification (`country`)
- `names_from` — the spreading dimension (`year`)
- `values_from` — the measure (no aggregation, just redistribution of `lifeExp`)

```{r}
wide_tidyverse <- gapminder |>
  select(country, year, lifeExp) |>
  pivot_wider(
    id_cols     = country,
    names_from  = year,
    values_from = lifeExp
  )

wide_tidyverse
```

Notice the column names (`1952`, `1957`, ..., `2007`) came directly from the distinct values in the `year` column which is the defining feature of any pivot.

### Approach 2: DuckDB `PIVOT` (without `USING`)

DuckDB's `PIVOT` command without a `USING` clause performs pure reshaping. This is equivalent to `pivot_wider()`. No aggregation is specified; values are simply redistributed. The output below only shows 6 rows.

```{r}
# Set up DuckDB and load the data
con_duck <- dbConnect(duckdb::duckdb())
dbWriteTable(con_duck, "gapminder", gapminder, overwrite = TRUE)

wide_duckdb <- dbGetQuery(con_duck, "
  PIVOT (SELECT country, year, lifeExp FROM gapminder)
  ON year
  USING first(lifeExp)
")

# Sort to match tidyverse output for comparison
wide_duckdb <- wide_duckdb |> arrange(country)

head(wide_duckdb, 6)
```

Note: even though `USING first(lifeExp)` looks like an aggregation, since each country-year combination is unique, `first()` always operates on exactly one value and simply returns it unchanged. The output is structurally identical to the `pivot_wider()` result, except that this output is not rounded like in Approach 1.

### Approach 3: Manual reshaping in DuckDB (`CASE WHEN`)

The manual approach makes every column fully explicit, with no built-in pivot mechanism involved at all.

```{r}
wide_manual <- dbGetQuery(con_duck, "
  SELECT
    country,
    MAX(CASE WHEN year = 1952 THEN lifeExp END) AS \"1952\",
    MAX(CASE WHEN year = 1957 THEN lifeExp END) AS \"1957\",
    MAX(CASE WHEN year = 1962 THEN lifeExp END) AS \"1962\",
    MAX(CASE WHEN year = 1967 THEN lifeExp END) AS \"1967\",
    MAX(CASE WHEN year = 1972 THEN lifeExp END) AS \"1972\",
    MAX(CASE WHEN year = 1977 THEN lifeExp END) AS \"1977\",
    MAX(CASE WHEN year = 1982 THEN lifeExp END) AS \"1982\",
    MAX(CASE WHEN year = 1987 THEN lifeExp END) AS \"1987\",
    MAX(CASE WHEN year = 1992 THEN lifeExp END) AS \"1992\",
    MAX(CASE WHEN year = 1997 THEN lifeExp END) AS \"1997\",
    MAX(CASE WHEN year = 2002 THEN lifeExp END) AS \"2002\",
    MAX(CASE WHEN year = 2007 THEN lifeExp END) AS \"2007\"
  FROM gapminder
  GROUP BY country
  ORDER BY country
")

head(wide_manual, 6)
```

`MAX()` is used here not to aggregate multiple values, but because `GROUP BY country` requires an aggregate function around non-grouped columns, and since each group contains exactly one row per year, `MAX()` simply returns that single value.

### Verifying all three outputs match

```{r}
# Align column types and order for comparison
tidy_sorted <- wide_tidyverse |>
  arrange(country) |>
  mutate(across(-country, as.numeric))

duck_sorted <- wide_duckdb |>
  arrange(country) |>
  rename_with(~ str_remove(.x, "^lifeExp_"), -country) |>
  mutate(across(-country, as.numeric))

manual_sorted <- wide_manual |>
  arrange(country) |>
  mutate(across(-country, as.numeric))

all.equal(tidy_sorted, duck_sorted,   check.attributes = FALSE)
all.equal(tidy_sorted, manual_sorted, check.attributes = FALSE)
```

All three approaches produce identical results — the same countries, the same years as column headers, and the same life expectancy values in every cell.

### Can `crosstab()` do pure reshaping?

Short answer: **not cleanly.** `crosstab()` always expects exactly one (row_id, category, value) triple per row of the source query, where the value has already been aggregated. In the pure reshaping case (one value per country-year), you *can* technically pass `crosstab()` a query that produces those triples althought it is a little more complicated than the previous options:

```{r}
con_local <- dbConnect(
  RPostgres::Postgres(),
  dbname   = "postgres",
  host     = "localhost",
  port     = 5432,
  user     = "sebastianpollack",
  password = "sup"
)
```

```{r}

# Requires a local PostgreSQL connection with tablefunc installed
dbExecute(con_local, "CREATE EXTENSION IF NOT EXISTS tablefunc;")
dbWriteTable(con_local, "gapminder", gapminder, overwrite = TRUE)

wide_crosstab <- dbGetQuery(con_local, '
  SELECT * FROM crosstab(
    \'SELECT country,
            year::text,
            "lifeExp"
     FROM gapminder
     ORDER BY country, year::text\',
    \'SELECT DISTINCT year::text FROM gapminder ORDER BY year::text\'
  ) AS ct(
    country text,
    "1952" float, "1957" float, "1962" float, "1967" float,
    "1972" float, "1977" float, "1982" float, "1987" float,
    "1992" float, "1997" float, "2002" float, "2007" float
  )
')

head(wide_crosstab,6)
```

This will produce the same output — but notice that `crosstab()` is doing exactly what `pivot_wider()` and the manual `CASE WHEN` do: retrieving one existing value per cell. `crosstab()`'s mandatory output-column declaration makes it more verbose and more fragile than `pivot_wider()` for a pure reshaping task.

---

## Part 2: Cross-Tabulation

### The data structure

Cross-tabulation is appropriate when **multiple source rows share the same (row-group, column-group) combination**, and those rows must be collapsed into a single cell via an aggregation function.

In our second example:

- **Rows:** continent
- **Columns:** year
- **Cells:** average life expectancy across all countries in that continent-year combination

Each continent contains multiple countries, so each (continent, year) pair has multiple source rows — aggregation is not optional, it's required.

```{r}
# How many countries per continent-year?
gapminder |>
  count(continent, year) |>
  pivot_wider(names_from = year, values_from = n)
```


### Approach 1: Manual (`CASE WHEN` + `AVG`)

```{r}
ct_manual <- dbGetQuery(con_duck, "
  SELECT
    continent,
    ROUND(AVG(CASE WHEN year = 1952 THEN lifeExp END), 2) AS \"1952\",
    ROUND(AVG(CASE WHEN year = 1957 THEN lifeExp END), 2) AS \"1957\",
    ROUND(AVG(CASE WHEN year = 1962 THEN lifeExp END), 2) AS \"1962\",
    ROUND(AVG(CASE WHEN year = 1967 THEN lifeExp END), 2) AS \"1967\",
    ROUND(AVG(CASE WHEN year = 1972 THEN lifeExp END), 2) AS \"1972\",
    ROUND(AVG(CASE WHEN year = 1977 THEN lifeExp END), 2) AS \"1977\",
    ROUND(AVG(CASE WHEN year = 1982 THEN lifeExp END), 2) AS \"1982\",
    ROUND(AVG(CASE WHEN year = 1987 THEN lifeExp END), 2) AS \"1987\",
    ROUND(AVG(CASE WHEN year = 1992 THEN lifeExp END), 2) AS \"1992\",
    ROUND(AVG(CASE WHEN year = 1997 THEN lifeExp END), 2) AS \"1997\",
    ROUND(AVG(CASE WHEN year = 2002 THEN lifeExp END), 2) AS \"2002\",
    ROUND(AVG(CASE WHEN year = 2007 THEN lifeExp END), 2) AS \"2007\"
  FROM gapminder
  GROUP BY continent
  ORDER BY continent
")

ct_manual
```

Here `AVG()` is taking the average of life expectancy per year for every continent.

### Approach 2: DuckDB `PIVOT` with `USING avg()`

```{r}
ct_duckdb <- dbGetQuery(con_duck, "
  PIVOT (SELECT continent, year, lifeExp FROM gapminder)
  ON year
  USING ROUND(AVG(lifeExp), 2)
  GROUP BY continent
  ORDER BY continent
")

ct_duckdb
```

The `USING avg(lifeExp)` clause is now performing averaging life expectancy given the specified conditions and formatting.

### Approach 3: `crosstab()` in PostgreSQL

```{r}
#| eval: false

dbWriteTable(con_local, "gapminder", gapminder, overwrite = TRUE)

dbGetQuery(con_local, "
  SELECT * FROM crosstab(
    'SELECT continent,
            year::text,
            ROUND(AVG(\"lifeExp\")::numeric, 2)
     FROM gapminder
     GROUP BY continent, year
     ORDER BY continent, year::text',
    'SELECT DISTINCT year::text FROM gapminder ORDER BY year::text'
  ) AS ct(
    continent text,
    \"1952\" numeric, \"1957\" numeric, \"1962\" numeric, \"1967\" numeric,
    \"1972\" numeric, \"1977\" numeric, \"1982\" numeric, \"1987\" numeric,
    \"1992\" numeric, \"1997\" numeric, \"2002\" numeric, \"2007\" numeric
  )
")
```

This is `crosstab()`'s main use case: the source query performs the `GROUP BY continent, year` aggregation first, producing one pre-aggregated (continent, year, avg_lifeExp) triple per row, which is the three-column structure `crosstab()` requires. It then reshapes those triples into the wide format.

### Verifying manual and DuckDB PIVOT match

```{r}
manual_sorted_ct <- ct_manual |> arrange(continent)
duck_sorted_ct   <- ct_duckdb |> arrange(continent)

all.equal(manual_sorted_ct, duck_sorted_ct, check.attributes = FALSE)
```

---

## Summary

| | Pure Reshaping | Cross-Tabulation |
|---|---|---|
| **Row count changes?** | No | Yes (fewer rows out) |
| **Aggregation required?** | No | Yes |
| **R tidyverse** | `pivot_wider()` | `pivot_wider()` + pre-aggregation |
| **DuckDB `PIVOT`** | `PIVOT ... ON ... USING first()` | `PIVOT ... ON ... USING avg()` |
| **Manual SQL** | `MAX(CASE WHEN ...)` + `GROUP BY` | `AVG(CASE WHEN ...)` + `GROUP BY` |
| **PostgreSQL `crosstab()`** | Works but awkward, no native mode for it | Natural fit, designed for pre-aggregated input |

```{r cleanup, include=FALSE}
dbDisconnect(con_duck, shutdown = TRUE)
```
