---
title: "Join Algorithms and Join Order in PostgreSQL"
format:
  html:
    toc: true
    toc-location: right
    code-fold: true
    code-summary: "Show code"
   
execute:
  warning: false
  message: false
---

## Introduction

When you write a multi-table `JOIN`, you specify *which* tables connect and *on what columns* — but you do **not** specify *how* the database should physically combine them, nor *in what order*. Those two decisions belong to the query optimizer, and they are the most important influences  on how fast a join runs.

This simulation isolates and measures two optimization decisions:

- **Join strategy** — the physical algorithm used to match rows between two tables. PostgreSQL has three: **Nested Loop**, **Hash Join**, and **Merge Join**.
For more information on these strategies, see the <a href="_site/joins.html">Joins page</a>.
- **Join order** — the sequence in which tables are brought into the join. Starting from heavily-filtered small tables keeps intermediate results tiny; starting from large unfiltered tables forces the engine to carry millions of rows through every step.

Normally the planner picks both automatically using table statistics and cost estimates.To *see* the effect of each choice, we override the planner with session-level `SET` commands, force one strategy and one order at a time, and time the result. This is purely a teaching device — in production you almost always let the planner decide.

We use the Stack Exchange `stats` dataset (the Cross Validated dump), which has four tables we can chain together: `votes`, `posts`, `users`, and `badges`.

### Setup

We connect to the database, build indexes on the join columns, and confirm the connection with a trivial query. Indexes matter here: a Nested Loop join is only competitive when the inner side can be probed through an index rather than scanned in full.

```{r}
#| label: setup
#| eval: false
library(DBI)
library(RPostgres)
library(purrr)
library(tidyverse)

con <- dbConnect(
  RPostgres::Postgres(),
  dbname = "stats",
  host   = "127.0.0.1",
  port   = 5432,
  user   = Sys.info()[["user"]]
)
```
```{r}

# Index the columns that appear in our JOIN ... ON clauses
dbExecute(con, "CREATE INDEX IF NOT EXISTS idx_votes_postid       ON votes(postid)")
dbExecute(con, "CREATE INDEX IF NOT EXISTS idx_posts_owneruserid  ON posts(owneruserid)")
dbExecute(con, "CREATE INDEX IF NOT EXISTS idx_badges_userid      ON badges(userid)")

dbGetQuery(con, "EXPLAIN ANALYZE SELECT COUNT(*) FROM votes")
```

## How the planner can be controlled

PostgreSQL exposes a set of session variables that let us switch individual planner behaviours on and off. These are the knobs we use throughout:

- `enable_nestloop`, `enable_hashjoin`, `enable_mergejoin` — turn each physical strategy on or off. By enabling exactly one at a time, we force the planner to use that strategy.
- `join_collapse_limit` / `from_collapse_limit` — when set to `1`, the planner is forbidden from reordering the tables and must join them in the literal order written in the `FROM` clause. This is what lets us dictate join *order*.
- `max_parallel_workers_per_gather` — set to `0` to turn off parallelism, so timing differences reflect the algorithm and not the number of workers.

A key idea: because `join_collapse_limit = 1` makes the engine respect the written order, the *same logical query* written with its tables in a different sequence becomes a different physical plan.

```{r}
#| label: planner-knobs
dbExecute(con, "SET max_parallel_workers_per_gather = 0")
dbExecute(con, "SET join_collapse_limit = 1")
dbExecute(con, "SET from_collapse_limit = 1")
```

## The two queries: good order vs. bad order

Both queries below answer the **exact same question** — count the rows produced by joining votes to their posts, those posts to their owners, and those owners to their badges, keeping only upvotes (`votetypeid = 2`) on well-scored posts (`score > 10`) owned by high-reputation users (`reputation > 1000`). They return identical results. The only difference is the order the tables are listed in.

- **Good order** leads with `votes` and `posts`, where the `WHERE` filters are most selective, so the very first join already works on a shrunken set of rows.
- **Bad order** leads with `badges` and `users`, the large tables whose filters don't bite until later, so early intermediate results stay huge.

```{r}
#| label: define-queries
# Good order: filtered/small tables first
sql_good_order <- "
  SELECT COUNT(*)
  FROM votes v
  JOIN posts  p ON v.postid = p.id
  JOIN users  u ON p.owneruserid = u.id
  JOIN badges b ON b.userid = u.id
  WHERE v.votetypeid = 2
    AND p.score > 10
    AND u.reputation > 1000
"

# Bad order: large/unfiltered tables first
sql_bad_order <- "
  SELECT COUNT(*)
  FROM badges b
  JOIN users  u ON b.userid = u.id
  JOIN posts  p ON p.owneruserid = u.id
  JOIN votes  v ON v.postid = p.id
  WHERE v.votetypeid = 2
    AND p.score > 10
    AND u.reputation > 1000
"
```

## A repeatable timing harness (warm cache)

To get a stable measurement we wrap `EXPLAIN ANALYZE` in a small helper. It runs one throwaway query first to load the relevant data pages into PostgreSQL's buffer cache (a **warm-cache** measurement), then runs the query `n` more times, scraping the `Execution Time` and `Planning Time` lines out of each plan.

`EXPLAIN ANALYZE` is the tool doing the real work here: it actually executes the query and reports how long each step took, as opposed to plain `EXPLAIN`, which only prints the planner's estimate.

```{r}
#| label: warm-harness
run_repeated_warm <- function(con, query, n = 10) {

  # Warm-up run: loads pages into cache, result discarded
  dbGetQuery(con, paste("EXPLAIN ANALYZE", query))

  # Timed trials
  map(1:n, function(i) {
    lines <- dbGetQuery(con, paste("EXPLAIN ANALYZE", query))[[1]]

    exec_time <- lines[grep("Execution Time", lines)] |>
      str_extract("[0-9.]+") |> as.numeric()

    plan_time <- lines[grep("Planning Time", lines)] |>
      str_extract("[0-9.]+") |> as.numeric()

    tibble(run = i, exec_ms = exec_time, plan_ms = plan_time)
  }) |>
  list_rbind()
}
```

### Running all six combinations

We now cross the two join orders with the three join strategies, forcing one strategy at a time by toggling the `enable_*` flags. That gives six timing distributions of ten runs each.

```{r}
#| label: warm-runs
# --- Nested Loop ---
dbExecute(con, "SET enable_nestloop  = on")
dbExecute(con, "SET enable_hashjoin  = off")
dbExecute(con, "SET enable_mergejoin = off")
results_nest_good <- run_repeated_warm(con, sql_good_order, 10) |>
  mutate(join_type = "Nested Loop", order = "good")
results_nest_bad  <- run_repeated_warm(con, sql_bad_order, 10) |>
  mutate(join_type = "Nested Loop", order = "bad")

# --- Hash Join ---
dbExecute(con, "SET enable_nestloop  = off")
dbExecute(con, "SET enable_hashjoin  = on")
dbExecute(con, "SET enable_mergejoin = off")
results_hash_good <- run_repeated_warm(con, sql_good_order, 10) |>
  mutate(join_type = "Hash Join", order = "good")
results_hash_bad  <- run_repeated_warm(con, sql_bad_order, 10) |>
  mutate(join_type = "Hash Join", order = "bad")

# --- Merge Join ---
dbExecute(con, "SET enable_nestloop  = off")
dbExecute(con, "SET enable_hashjoin  = off")
dbExecute(con, "SET enable_mergejoin = on")
results_merge_good <- run_repeated_warm(con, sql_good_order, 10) |>
  mutate(join_type = "Merge Join", order = "good")
results_merge_bad  <- run_repeated_warm(con, sql_bad_order, 10) |>
  mutate(join_type = "Merge Join", order = "bad")
```

```{r}
#| label: warm-plot
#| fig-cap: "Warm-cache execution time across the six strategy × order combinations."
bind_rows(
  results_nest_good, results_nest_bad,
  results_hash_good, results_hash_bad,
  results_merge_good, results_merge_bad
) |>
  ggplot(aes(x = join_type, y = exec_ms, color = order)) +
  geom_boxplot() +
  labs(
    title    = "Effect of join strategy on execution time",
    subtitle = "4-table join: votes, posts, users, badges",
    x        = "Join Strategy",
    y        = "Execution time (ms)"
  ) +
  theme_minimal()
```

**What the plot shows.** Each strategy responds differently to a bad join order, and that is the whole point. A **Nested Loop** is the most order-sensitive: for every row from the outer side it probes the inner side, so if the outer side starts large (bad order) the probe count explodes. A **Hash Join** builds a hash table on one input and streams the other past it, which makes it far more forgiving of a bad starting order. A **Merge Join** requires both inputs sorted on the join key, so its cost is dominated by sorting and it sits somewhere in between. In short: a good join order helps every strategy, but it *rescues* the Nested Loop.

## Sanity check: the two orders return the same answer

Before trusting the timings, we confirm that "good" and "bad" really are the same query logically — same count — and we list the indexes currently in place.

```{r}
#| label: verify-equal
dbGetQuery(con, sql_good_order)
dbGetQuery(con, sql_bad_order)

dbGetQuery(con, "
  SELECT tablename, indexname
  FROM pg_indexes
  WHERE schemaname = 'public'
  ORDER BY tablename
")
```

## A note on cold cache

Every measurement above is **warm-cache**: the warm-up run loads the
relevant pages into PostgreSQL's buffer pool, so the timed runs reflect
the cost of the join algorithm rather than the cost of fetching data from
disk. This is the right thing to measure for a query that runs often
enough to keep its pages resident in memory.

A **cold-cache** test — closer to a first-of-the-day query, or an
analytics query over tables too large to stay in RAM — would be a useful
counterpart, but it is genuinely hard to do faithfully here. A real cold
measurement has to evict data from two separate layers: PostgreSQL's own
`shared_buffers`, and the operating system's page cache underneath it.
Neither is reachable from inside a database session. `DISCARD ALL` only
resets session state (planner settings, prepared statements, temp tables)
and leaves both caches fully populated; flushing the OS page cache
requires root access we don't have in this environment. Reconnecting on
each iteration looks colder but measures essentially the same warm pages.

So rather than report numbers from a harness that can't actually clear the
cache, we'll state the expectation instead: with a true cold cache the
**order penalty should widen, not shrink**. A bad join order carries more
rows through each step and therefore touches more pages; when those pages
must be read from disk instead of memory, the extra I/O falls hardest on
exactly the plan that was already doing the most work — the large-table-first
Nested Loop. The warm-cache results below are thus a *conservative* picture
of how much join order matters; on cold data the gap between good and bad
order would only grow.

## Why the planner makes these choices: estimated vs. actual rows

The planner doesn't time queries — it *estimates* how many rows each step will emit and picks the cheapest plan based on those estimates. When the estimates are wrong, the plan is wrong. The function below pulls every node out of a plan and compares the planner's **estimated** row count to the **actual** row count, computing a **q-error**: the ratio of the larger to the smaller. A q-error near 1 means the estimate was accurate; a large q-error means the planner was badly misled at that node.

```{r}
#| label: estimate-harness
extract_estimates <- function(con, sql, label) {
  plan <- dbGetQuery(con, paste("EXPLAIN ANALYZE", sql))[[1]]

  # Keep only nodes that report actual timing
  node_lines <- plan[grep("actual time", plan)]

  map(node_lines, function(line) {
    operation <- str_extract(line,
      "(Nested Loop|Hash Join|Merge Join|Seq Scan|Index Only Scan|Index Scan|Aggregate|Gather)")

    estimated <- str_extract(line, "rows=([0-9]+)\\s+width") |>
      str_extract("[0-9]+") |> as.numeric()

    actual <- str_extract(line, "actual time=[0-9.]+\\.\\.[0-9.]+ rows=([0-9]+)") |>
      str_extract("[0-9]+$") |> as.numeric()

    loops <- str_extract(line, "loops=([0-9]+)") |>
      str_extract("[0-9]+") |> as.numeric()

    tibble(
      label        = label,
      operation    = operation,
      estimated    = estimated,
      actual       = actual,
      loops        = loops,
      actual_total = actual * loops,
      q_error      = round(pmax(estimated / actual, actual / estimated), 1)
    )
  }) |>
  list_rbind() |>
  filter(!is.na(operation), !is.na(estimated), !is.na(actual))
}
```

```{r}
#| label: estimate-runs
dbExecute(con, "SET max_parallel_workers_per_gather = 0")

# Forced orders: planner not allowed to reorder
dbExecute(con, "SET join_collapse_limit = 1")
good_estimates <- extract_estimates(con, sql_good_order, "forced good order")
bad_estimates  <- extract_estimates(con, sql_bad_order,  "forced bad order")

# Auto-optimized: planner free to choose its own order
dbExecute(con, "SET join_collapse_limit = 8")
auto_estimates <- extract_estimates(con, sql_good_order, "auto-optimized")

# Reset session to normal behaviour
dbExecute(con, "SET max_parallel_workers_per_gather = 4")
dbExecute(con, "SET join_collapse_limit = 8")

all_estimates <- bind_rows(good_estimates, bad_estimates, auto_estimates)
all_estimates
```

**Reading this table.** The `auto-optimized` rows show what the planner does when left alone (`join_collapse_limit = 8` lets it reorder up to eight tables freely). Comparing q-errors across the three labels shows where estimation breaks down: nodes deep in the bad-order plan tend to carry the largest q-errors, because each early misestimate compounds as it feeds the next join. This is the mechanism behind the timing differences — a bad order isn't slow by fiat, it's slow because the planner's row estimates drift further from reality at every step, and the `actual_total` (rows × loops) balloons.

```{r}
#| label: cleanup
dbDisconnect(con)
```

## Summary

|                                | Nested Loop                          | Hash Join                              | Merge Join                          |
| ------------------------------ | ------------------------------------ | -------------------------------------- | ----------------------------------- |
| **How it matches rows**        | Probe inner table once per outer row | Build hash on one side, stream other   | Sort both sides, walk in lockstep   |
| **Needs an index?**            | Strongly benefits from inner index   | No                                      | Benefits from pre-sorted inputs     |
| **Sensitivity to join order**  | Very high                            | Low                                     | Moderate                            |
| **Best when**                  | Outer side is small after filtering  | Inputs are large and unsorted          | Inputs already sorted on the key    |

Two takeaways. First, **join order matters most for Nested Loops**: leading with your most-filtered tables keeps the outer row count — and therefore the probe count — small. Second, the planner's choices flow directly from its **row estimates**, so accurate statistics (and thus low q-error) let the optimizer pick the right strategy and order on its own. Forcing these settings is useful for *understanding* the engine, but in practice the lesson is to keep statistics fresh and write selective filters, then trust the planner.
