---
title: "Native DuckDB vs Direct Parquet Querying Performance"
format:
  html:
    toc: true
    toc-depth: 2
    toc-title: "Contents"
    code-fold: true
    code-summary: "Show code"
    self-contained: true
execute:
  warning: false
  message: false
bibliography: references.bib
suppress-bibliography: true
---

## Introduction

This page reports the results of a controlled benchmarking experiment comparing query performance between two approaches to storing and querying the same data: a **native DuckDB database** and **direct Parquet files** queried via `read_parquet()`. Both formats contain identical data drawn from three California city subsets of the [Stanford Open Policing Project](https://openpolicing.stanford.edu/) [[@openpolicing2023]](references.qmd) — Los Angeles, San Francisco, and San Jose — sourced from the same MySQL database. The only difference between the two approaches is the storage format itself.

For background on Parquet's internal structure and how DuckDB reads files directly, see the [Parquet & DuckDB reference page](parquet.qmd).

Six query types are benchmarked to represent a range of workloads, from simple counts to cross-table aggregations:

| Query | Description |
|---|---|
| **Q1** | `COUNT(*)` — total row count across a single city |
| **Q2** | `GROUP BY subject_race` — aggregation on one column, single city |
| **Q3** | `GROUP BY type` — aggregation on a different column, single city |
| **Q4** | String filter — `ILIKE` pattern match on a string column |
| **Q5** | Cross-table `GROUP BY subject_race` — aggregation across all three cities |

Each timing result is the **mean of 20 timed runs**, preceded by 3 warmup runs. A fresh DuckDB connection is opened and closed for every trial to avoid connection-level caching effects.

### Statistical Significance

Differences between native DuckDB and direct Parquet are tested using a **two-sided Wilcoxon rank-sum test** at α = 0.05. This non-parametric test is appropriate here because query times are non-negative, right-skewed, and not guaranteed to follow a normal distribution. A result marked **Yes*** indicates the observed difference in timing distributions is statistically unlikely to have occurred by chance, and we can conclude the two approaches differ meaningfully in performance for that query. A result marked **No** means the difference is not statistically distinguishable from random variation.

---

## Setup

```{r}
#| label: setup

library(arrow)
library(duckdb)
library(DBI)
library(dplyr)
library(tidyr)
library(ggplot2)
library(here)
library(RMySQL)
```

---

## Step 1: Build Storage Formats {.unnumbered}

This step can be done in many approaches. For this guide, a connection with MySQL is setup, then three tables from the entire SQL database are used to write to its own Parquet file. Then a native DuckDB database is built with each city (Parquet file) as its own table. This only needs to run once so set `eval: false` after the databases are built.

```{r}
#| label: build-databases
#| eval: false

Sys.setenv(MARIADB_TLS_DISABLE_PEER_VERIFICATION = 1)

con_traffic <- DBI::dbConnect(
  RMySQL::MySQL(),
  dbname   = "traffic",
  host     = Sys.getenv("TRAFFIC_HOST"),
  user     = Sys.getenv("TRAFFIC_USER"),
  password = Sys.getenv("TRAFFIC_PWD")
)

tables     <- c("ca_los_angeles_2020_04_01",
                "ca_san_francisco_2020_04_01",
                "ca_san_jose_2020_04_01")
short_names <- c("la", "sf", "sj")

# ── Write one Parquet file per city ──
dir.create(here::here("parquet_cities"), showWarnings = FALSE)

for (i in seq_along(tables)) {
  cat("Reading", tables[i], "from MySQL...\n")
  df      <- dbReadTable(con_traffic, tables[i])
  outfile <- here::here("parquet_cities", paste0(short_names[i], ".parquet"))
  write_parquet(df, outfile)
  cat("  Written:", outfile, "\n")
  rm(df); gc()
}

# ── Build native DuckDB — one table per city ──
con_native <- dbConnect(duckdb(), here::here("traffic_native.duckdb"))

for (i in seq_along(tables)) {
  cat("Loading", tables[i], "into DuckDB...\n")
  df <- dbReadTable(con_traffic, tables[i])
  dbWriteTable(con_native, short_names[i], df, overwrite = TRUE)
  cat("  Rows written:", nrow(df), "\n")
  rm(df); gc()
}

cat("\nTables in native DuckDB:\n")
print(dbListTables(con_native))

dbDisconnect(con_native, shutdown = TRUE)
dbDisconnect(con_traffic)
cat("Done.\n")
```

---

## Step 2: Storage Size Comparison

One concrete advantage of direct Parquet over native DuckDB is storage footprint and size. Because Parquet files are written with columnar compression applied at the column level, they are typically substantially smaller than an equivalent DuckDB database, which stores data in its own internal format optimized for query performance rather than disk compactness.

```{r}
#| label: storage-size

parquet_files <- list.files(here::here("parquet_cities"),
                             pattern = "\\.parquet$", full.names = TRUE)
parquet_size_mb <- round(sum(file.size(parquet_files)) / 1e6, 2)
parquet_size_gb <- round(parquet_size_mb / 1000, 3)

duckdb_size_mb <- round(file.size(here::here("traffic_native.duckdb")) / 1e6, 2)
duckdb_size_gb <- round(duckdb_size_mb / 1000, 3)

ratio <- round(duckdb_size_mb / parquet_size_mb, 2)

size_df <- data.frame(
  Format        = c("Direct Parquet (3 files)", "Native DuckDB (1 file)"),
  `Size (MB)`   = c(parquet_size_mb, duckdb_size_mb),
  `Size (GB)`   = c(parquet_size_gb, duckdb_size_gb),
  `Files`       = c(3, 1),
  `Ratio vs Parquet` = c(1.00, ratio)
)

knitr::kable(size_df, align = "r",
  caption = "Storage size comparison: direct Parquet vs native DuckDB")
```

Beyond storage size, direct Parquet also avoids the ingestion step entirely. The data can be queried immediately after the Parquet files are written, with no loading or database construction required. The trade-off is query performance, which the remainder of this page explores.

---

## Step 3: Benchmark

This section contains the code for creating the queries and obtaining the mean times for each query. In this code, 5 queries were performed and are meant to represent varying forms of queries, from counting rows to cross-table queries.

```{r}
#| label: define-queries

la_parquet <- here::here("parquet_cities", "la.parquet")
sf_parquet <- here::here("parquet_cities", "sf.parquet")
sj_parquet <- here::here("parquet_cities", "sj.parquet")

queries <- list(

  Q1 = list(
    label  = "COUNT(*) — LA only",
    direct = sprintf("SELECT COUNT(*) FROM read_parquet('%s')", la_parquet),
    native = "SELECT COUNT(*) FROM la"
  ),

  Q2 = list(
    label  = "GROUP BY subject_race — LA only",
    direct = sprintf(
      "SELECT subject_race, COUNT(*) AS n FROM read_parquet('%s') GROUP BY subject_race ORDER BY n DESC",
      la_parquet),
    native = "SELECT subject_race, COUNT(*) AS n FROM la GROUP BY subject_race ORDER BY n DESC"
  ),

  Q3 = list(
    label  = "GROUP BY type — LA only",
    direct = sprintf(
      "SELECT type, COUNT(*) AS n FROM read_parquet('%s') GROUP BY type ORDER BY n DESC",
      la_parquet),
    native = "SELECT type, COUNT(*) AS n FROM la GROUP BY type ORDER BY n DESC"
  ),

  Q4 = list(
    label  = "String filter ILIKE — LA only",
    direct = sprintf(
      "SELECT COUNT(*) FROM read_parquet('%s') WHERE raw_descent_description ILIKE '%%hispanic%%'",
      la_parquet),
    native = "SELECT COUNT(*) FROM la WHERE raw_descent_description ILIKE '%hispanic%'"
  ),

  Q5 = list(
    label  = "Cross-table GROUP BY subject_race — all three cities",
    direct = sprintf(
      "SELECT subject_race, COUNT(*) AS n FROM (
         SELECT subject_race FROM read_parquet('%s')
         UNION ALL
         SELECT subject_race FROM read_parquet('%s')
         UNION ALL
         SELECT subject_race FROM read_parquet('%s')
       ) GROUP BY subject_race ORDER BY n DESC",
      la_parquet, sf_parquet, sj_parquet),
    native = "SELECT subject_race, COUNT(*) AS n FROM (
         SELECT subject_race FROM la
         UNION ALL
         SELECT subject_race FROM sf
         UNION ALL
         SELECT subject_race FROM sj
       ) GROUP BY subject_race ORDER BY n DESC"
  )
)
```

```{r}
#| label: benchmark

fresh_direct <- function() dbConnect(duckdb())
fresh_native <- function() dbConnect(duckdb(), here::here("traffic_native.duckdb"))

n_warmup <- 3
n_trials <- 20

run_trials <- function(con_fn, sql, n_warmup, n_trials) {
  for (i in seq_len(n_warmup)) {
    con <- con_fn()
    try(dbGetQuery(con, sql), silent = TRUE)
    try(dbDisconnect(con, shutdown = TRUE), silent = TRUE)
    gc()
  }
  times_ms <- numeric(n_trials)
  for (i in seq_len(n_trials)) {
    con <- con_fn()
    tryCatch({
      times_ms[i] <- system.time(dbGetQuery(con, sql))["elapsed"] * 1000
    }, error = function(e) {
      times_ms[i] <<- NA
    }, finally = {
      try(dbDisconnect(con, shutdown = TRUE), silent = TRUE)
    })
    gc()
  }
  times_ms
}

all_trials <- data.frame(
  query_id   = character(),
  label      = character(),
  format     = character(),
  trial      = integer(),
  time_ms    = numeric()
)

for (qid in names(queries)) {
  q <- queries[[qid]]
  cat("\nRunning", qid, "—", q$label, "\n")

  cat("  Native...\n")
  nt <- run_trials(fresh_native, q$native, n_warmup, n_trials)
  cat("  Native mean:", round(mean(nt, na.rm = TRUE), 1), "ms\n")

  cat("  Direct...\n")
  dt <- run_trials(fresh_direct, q$direct, n_warmup, n_trials)
  cat("  Direct mean:", round(mean(dt, na.rm = TRUE), 1), "ms\n")

  all_trials <- rbind(all_trials,
    data.frame(query_id = qid, label = q$label, format = "native",
               trial = seq_len(n_trials), time_ms = round(nt, 3)),
    data.frame(query_id = qid, label = q$label, format = "direct",
               trial = seq_len(n_trials), time_ms = round(dt, 3))
  )
}
```

---

## Results

### Summary Table

```{r}
#| label: results-table

summary_df <- all_trials %>%
  group_by(query_id, label, format) %>%
  summarise(
    mean_ms = round(mean(time_ms, na.rm = TRUE), 1),
    .groups   = "drop"
  ) %>%
  pivot_wider(names_from = format, values_from = mean_ms) %>%
  mutate(
    diff_ms = round(direct - native, 1),
    ratio   = round(direct / native, 2),
    faster  = ifelse(native < direct, "Native", "Direct Parquet")
  )

# Add significance test
sig_col <- mapply(function(qid) {
  nt <- all_trials %>% filter(query_id == qid, format == "native") %>% pull(time_ms)
  dt <- all_trials %>% filter(query_id == qid, format == "direct") %>% pull(time_ms)
  p  <- tryCatch(
    wilcox.test(nt, dt, exact = FALSE)$p.value,
    error = function(e) NA
  )
  ifelse(!is.na(p) & p < 0.05, "Yes *", "No")
}, summary_df$query_id)

summary_df$significant <- sig_col

knitr::kable(
  summary_df %>%
    select(label, native, direct, diff_ms, ratio, faster, significant),
  col.names = c("Query", "Native (ms)", "Direct (ms)",
                "Diff (ms)", "Ratio (D/N)", "Faster", "Sig. (p < 0.05)"),
  align = c("l", "r", "r", "r", "r", "l", "c"),
  caption = "Mean query time: native DuckDB vs direct Parquet. Ratio > 1 means direct Parquet is slower. Significance assessed via Wilcoxon rank-sum test."
)
```

### Visual Representation

The chart below shows the direct/native ratio for each query type. A ratio above 1 (dashed line) means direct Parquet was slower than native DuckDB for that query; below 1 means direct Parquet was faster.

```{r}
#| label: lollipop-chart
#| fig-cap: "Direct-to-native query time ratio by query type. Ratio > 1 indicates native DuckDB is faster; ratio < 1 indicates direct Parquet is faster."
#| fig-width: 9
#| fig-height: 5

plot_df <- summary_df %>%
  mutate(
    label     = factor(label, levels = rev(label)),
    color_grp = ifelse(ratio > 1, "Native faster", "Direct faster")
  )

ggplot(plot_df, aes(x = ratio, y = label, colour = color_grp)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "#94a3b8", linewidth = 0.8) +
  geom_segment(aes(x = 1, xend = ratio, y = label, yend = label),
               linewidth = 1.1) +
  geom_point(size = 5) +
  geom_text(aes(label = sprintf("%.2f×", ratio)),
            hjust = ifelse(plot_df$ratio > 1, -0.3, 1.3),
            size = 3.5, fontface = "bold") +
  scale_colour_manual(values = c("Native faster" = "#185FA5",
                                 "Direct faster" = "#0F6E56")) +
  scale_x_continuous(
    limits = c(
      min(0.5, min(plot_df$ratio) * 0.85),
      max(1.5, max(plot_df$ratio) * 1.15)
    ),
    breaks = seq(0, 10, by = 0.5)
  ) +
  labs(
    x      = "Ratio (Direct / Native)",
    y      = NULL,
    colour = NULL,
    title  = "Direct Parquet vs Native DuckDB: Query Time Ratio",
    subtitle = "Dashed line at 1.0 = equal performance"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position  = "top",
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank()
  )
```

---

## Conclusion

The results show a distinguishable difference that native DuckDB offers faster and more consistent query performance across most or all of the query types. Direct Parquet offers the advantage that files can be queried immediately after writing and occupy significantly less disk space.

The results can be understood based on the knowledge of how DuckDB querying and direct Parquet querying functions. When data is ingested into a native DuckDB database, DuckDB builds and maintains its own internal columnar storage format, statistics, and metadata that are optimized specifically for its query engine. When a query runs against a native table, DuckDB can exploit these internals directly, for example, reading only the pages it needs, using pre-computed column statistics, and operating entirely within a single file with no coordination overhead. Direct Parquet querying, by contrast, requires DuckDB to open each file, parse its footer, decompress column chunks, and assemble results from separate sources on every query execution. Even though Parquet is also columnar and compressed, this per-query parsing and file-handling overhead adds time that the native format avoids entirely. 

For workflows involving repeated analytical queries over stable data, native DuckDB is the stronger choice. For workflows where storage efficiency, portability, or avoiding ingestion latency matters more than raw query speed, direct Parquet is competitive and in some cases faster.
